---
title: "Combine FASTA files and align sequences"
format:
  html:
    toc: true
    toc-depth: 3
---

## Overview

`combineFASTA()` merges multiple `FASTA` files that all contain the **same locus** into a single file. It belongs to the sequence retrieval and preprocessing stage of a `catGenes` workflow. Therefore, it is a function you must run per locus *before* entering the actual alignment and concatenation workflow, not a step that mixes different genes together.

A `FASTA` file for a given locus commonly ends up split across several files because it came from more than one source or batch, for example:

- a search with `mineSeq()` or `mineTaxa()` split across more than one run or accession table
- the same locus retrieved both directly (`mineSeq()`/`mineTaxa()`) **and** extracted from whole plastid or mitochondrial genomes (`minePlastome()`/`mineMitochondrion()`)
- taxa fished out of a previous alignment for that locus with `fishSeqs()`, to be merged with other sequences of the same locus

In `catGenes`, this preprocessing step is paired with one more function before you move on to the alignment and concatenation workflow proper:

- `combineFASTA()` to combine multiple `FASTA` files of the *same locus* into a single file
- `alignSeqs()` to perform automated multiple sequence alignment on each locus file

This article explains when to combine `FASTA` files, how to run alignments, how to choose output formats, and how to prepare aligned files for later use in `catGenes`.

## Typical workflow

A common sequence-processing workflow is:

**Sequence retrieval and preprocessing** (per locus)

1. retrieve or mine sequences for that locus, possibly from more than one source
2. save them as `FASTA` files
3. if the locus ended up split across more than one file, combine them into a single `FASTA` file with `combineFASTA()`

**Alignment and concatenation workflow**

4. align the sequences for each locus
5. save aligned output in `NEXUS`, `FASTA`, or `PHYLIP`
6. load the alignments into R for downstream concatenation

Not every project requires the `combineFASTA()` step. But if each locus is already in a single `FASTA` file, go straight to `alignSeqs()`.

## When to use `combineFASTA()`

Use `combineFASTA()` when the **same locus** is split across multiple `FASTA` files, for example because:

- a search with `mineSeq()` or `mineTaxa()` was run more than once for that locus
- you want to merge separate downloads of the same locus into one file before alignment
- **the same locus was mined from two different sources**; for example, some accessions were retrieved as individual gene sequences with `mineSeq()` or `mineTaxa()`, while others were only available as whole plastid genomes and had to be extracted for that gene with `minePlastome()` (see [Combining the same locus mined from different sources](#combining-the-same-locus-mined-from-different-sources) below)
- you fished taxa for a locus out of a previous alignment with `fishSeqs()` and want to merge that output with other sequences of the same locus

`combineFASTA()` should **not** be used to merge files belonging to different genes into one file; each locus should stay in its own `FASTA` file going into `alignSeqs()`, so that alignment and, later, concatenation still operate on one locus at a time.

If your sequences are already in one `FASTA` file per locus, you may go directly to `alignSeqs()`.

## Basic usage of `combineFASTA()`

A simple workflow looks like this:

```{r, eval=FALSE}
library(catGenes)

result <- combineFASTA(
  input.files = c("matK_batch1.fasta", "matK_batch2.fasta", "matK_batch3.fasta")
)
```

This reads three `FASTA` files that all contain `matK` sequences (for instance three separate `mineSeq()` runs against different accession tables) combines all of them into a single object, and by default saves the result to disk.

## Choosing the output file name
You can specify a custom file name for the combined output, and the input file can come from different folder directories.
```{r, eval=FALSE}
result <- combineFASTA(
  input.files = c("data_mineSeqs/matK_part1.fasta", "data_mineTaxa/matK_part2.fasta"),
  output.file = "matK_combined.fasta"
)
```
Name the output after the locus itself — this keeps it clear that the combined file, like its inputs, holds a single gene.

## Returning the combined sequences without saving
If you want to inspect the combined sequences in R without immediately writing them to disk, set `save = FALSE`.

```{r, eval=FALSE}
result <- combineFASTA(
  input.files = c("file1.fasta", "file2.fasta"),
  save = FALSE,
  verbose = TRUE
)
```

In this case, the function returns the combined sequence object and summary information without creating an output file.

## Inspecting the result of combineFASTA()
The result returned by `combineFASTA()` includes:

- sequences: a DNAbin object containing all combined sequences
- summary: a data frame with summary information
- output_path: the output file path when saving is enabled

## Notes on duplicate sequences
`combineFASTA()` does not remove duplicate sequences. It simply combines all sequences from the files you specify.

This means that:
- repeated accessions remain in the combined file
- duplicate taxa are not filtered automatically
- downstream alignment and inspection may still be needed before concatenation

This behavior is useful because it preserves the original contents of the input files.

## When to use `alignSeqs()`
Use `alignSeqs()` when you have one or more `FASTA` files containing unaligned sequences and want to generate aligned outputs for downstream phylogenetic workflows.

The function uses the msa package and supports:

- ClustalW
- Muscle

It can write aligned results in:

- NEXUS
- FASTA
- PHYLIP

## Basic alignment workflow
A minimal alignment workflow looks like this:
```{r, eval=FALSE}
alignSeqs(
  filepath = "path_to_fasta_files",
  method = "ClustalW",
  format = "NEXUS"
)
```

This reads one or more `FASTA` files from the specified directory, aligns them, and writes the aligned output in `NEXUS` format.

## Choosing an alignment method
Currently, `alignSeqs()` supports two main alignment algorithms:

- ClustalW
- Muscle

For example, using `ClustalW`:
```{r, eval=FALSE}
alignSeqs(
  filepath = "path_to_fasta_files",
  method = "ClustalW",
  format = "NEXUS"
)
```

Or using `Muscle`:
```{r, eval=FALSE}
alignSeqs(
  filepath = "path_to_fasta_files",
  method = "Muscle",
  format = "NEXUS"
)
```

Both methods are widely used, and the best choice may depend on the characteristics of the dataset.

## Choosing the output format
The argument format determines how the aligned files are written. Supported output formats are:

- NEXUS
- FASTA
- PHYLIP

For example, to write aligned files in `FASTA` format:

```{r, eval=FALSE}
alignSeqs(
  filepath = "path_to_fasta_files",
  method = "ClustalW",
  format = "FASTA"
)
```

Or in `PHYLIP` format:

```{r, eval=FALSE}
alignSeqs(
  filepath = "path_to_fasta_files",
  method = "Muscle",
  format = "PHYLIP"
)
```

In most `catGenes` concatenation workflows, `NEXUS` is the most convenient format.

## Adjusting the gap opening penalty
The argument `gapOpening` controls the gap opening penalty used by the alignment algorithm.
```{r, eval=FALSE}
alignSeqs(
  filepath = "path_to_fasta_files",
  method = "ClustalW",
  gapOpening = "default",
  format = "NEXUS"
)
```

In many cases, the default setting is adequate. However, if your dataset requires specific tuning, you can adjust the value depending on the selected alignment method and the behavior you want.

## A complete sequence-processing example
A typical workflow for a single locus that was retrieved in two separate batches might look like this — for example, `ITS` accessions gathered in one project phase, with more `ITS` accessions added later from a second accession table.

Step 1. Retrieve the same locus in two batches
```{r, eval=FALSE}
batch1 <- mineSeq(
  inputdf = accession_table_batch1,
  gb.colnames = "ITS",
  save = TRUE,
  filename = "GenBank_seqs_batch1",
  dir = "RESULTS_mineSeq"
)

batch2 <- mineSeq(
  inputdf = accession_table_batch2,
  gb.colnames = "ITS",
  save = TRUE,
  filename = "GenBank_seqs_batch2",
  dir = "RESULTS_mineSeq"
)
```
This writes two separate files, both containing `ITS` sequences: `RESULTS_mineSeq/09Mar2026/GenBank_seqs_batch1_ITS.fasta` and `RESULTS_mineSeq/09Mar2026/GenBank_seqs_batch2_ITS.fasta`.

Step 2. Combine both batches into a single locus file
```{r, eval=FALSE}
combined <- combineFASTA(
  input.files = c(
    "RESULTS_mineSeq/09Mar2026/GenBank_seqs_batch1_ITS.fasta",
    "RESULTS_mineSeq/09Mar2026/GenBank_seqs_batch2_ITS.fasta"
  ),
  output.file = "ITS.fasta",
  dir = "RESULTS_combineFASTA"
)
```

Step 3. Align the combined locus
```{r, eval=FALSE}
alignSeqs(
  filepath = "RESULTS_combineFASTA/09Mar2026",
  method = "ClustalW",
  format = "NEXUS",
  dir = "RESULTS_alignSeqs"
)
```

This produces an aligned `ITS` file, ready for loading into R and later concatenation with the other loci in your dataset.

## Combining the same locus mined from different sources

A very common scenario is needing the *same* locus from more than one source at once:

- some accessions were submitted to GenBank as individual gene sequences, and are retrieved directly with `mineSeq()` (from a table of accessions) or `mineTaxa()` (from a taxonomic query)
- other accessions are only available as whole plastid genomes, so the same locus has to be extracted from each plastome with `minePlastome()`
- taxa were fished out of a previous alignment for that same locus with `fishSeqs()` (see [Fish and rename taxa across multiple alignments](/articles/fish_and_rename_taxa.qmd)) and need to be added to the rest

Each of these routes ends up producing its own separate `FASTA` file for the *same* gene. Before alignment, you want a single combined file per locus that includes accessions from all sources — this is exactly what `combineFASTA()` is for.

Step 1. Mine specific genes directly from individual GenBank accessions
```{r, eval=FALSE}
seqs <- mineSeq(
  inputdf = my_accession_table,
  gb.colnames = c("matK", "rbcL"),
  save = TRUE,
  filename = "GenBank_seqs",
  dir = "RESULTS_mineSeq"
)
```
This writes one file per gene, for example `RESULTS_mineSeq/09Mar2026/GenBank_seqs_matK.fasta` and `RESULTS_mineSeq/09Mar2026/GenBank_seqs_rbcL.fasta`.

Step 2. Mine the same genes from whole plastome accessions
```{r, eval=FALSE}
minePlastome(
  genbank = data_plastomes$Plastome,
  taxon = data_plastomes$Species,
  voucher = data_plastomes$Voucher,
  CDS = TRUE,
  genes = c("matK", "rbcL"),
  verbose = TRUE,
  dir = "RESULTS_minePlastome"
)
```
This also writes one file per gene, named after the gene itself: `RESULTS_minePlastome/09Mar2026/matK_from_plastome.fasta` and `RESULTS_minePlastome/09Mar2026/rbcL_from_plastome.fasta`.

Step 3. (optional) Add taxa fished from a previous locus alignment
Suppose you also want to add `matK` sequences for a few extra taxa that you fished out of someone else's published multi-locus alignments with `fishSeqs()`:
```{r, eval=FALSE}
fishSeqs(
  filepath = "Loci_alignment_external_dataset",
  taxa = c("Genus_speciesA", "Genus_speciesB"),
  dir = "RESULTS_fishSeqs"
)
```
`fishSeqs()` keeps each output file named after its original input file, so fishing from a `matK.fasta` alignment writes `RESULTS_fishSeqs/09Mar2026/matK.fasta` — again a single-locus file, this time with just the taxa you asked for.

Step 4. Combine all sources into a single FASTA file per locus
```{r, eval=FALSE}
combineFASTA(
  input.files = c(
    "RESULTS_mineSeq/09Mar2026/GenBank_seqs_matK.fasta",
    "RESULTS_minePlastome/09Mar2026/matK_from_plastome.fasta",
    "RESULTS_fishSeqs/09Mar2026/matK.fasta"
  ),
  output.file = "matK.fasta",
  dir = "RESULTS_combineFASTA"
)

combineFASTA(
  input.files = c(
    "RESULTS_mineSeq/09Mar2026/GenBank_seqs_rbcL.fasta",
    "RESULTS_minePlastome/09Mar2026/rbcL_from_plastome.fasta"
  ),
  output.file = "rbcL.fasta",
  dir = "RESULTS_combineFASTA"
)
```
`RESULTS_combineFASTA/09Mar2026/` now contains one `FASTA` file per locus (`matK.fasta`, `rbcL.fasta`), each combining accessions from every source mined for that gene — never a mix of different genes in one file.

Step 5. Align each combined locus
```{r, eval=FALSE}
alignSeqs(
  filepath = "RESULTS_combineFASTA/09Mar2026",
  method = "ClustalW",
  format = "NEXUS",
  dir = "RESULTS_alignSeqs"
)
```

Because `combineFASTA()` does not remove duplicates (see [Notes on duplicate sequences](#notes-on-duplicate-sequences) above), inspect each combined file before alignment in the rare case where the same accession was mined through both routes.

## Aligning multiple FASTA files from one folder
If a directory contains several `FASTA` files, `alignSeqs()` can process them together.

```{r, eval=FALSE}
alignSeqs(
  filepath = "RESULTS_mineSeq/09Mar2026",
  method = "Muscle",
  format = "NEXUS",
  dir = "RESULTS_alignSeqs"
)
```

This is useful when each file represents a different locus and all need to be aligned before loading into R.

## Output directories
By default, aligned files are written to a directory such as:

```{r, eval=FALSE}
RESULTS_alignSeqs/09Mar2026/
```

Similarly, combined `FASTA` outputs may be written to:

```{r, eval=FALSE}
RESULTS_combineFASTA/09Mar2026/
```

This date-based structure helps keep results organized across independent runs.

## File naming after alignment
If filename is not specified in `alignSeqs()`, output files are usually named based on the original input file names, with an added identifier such as aligned.
This makes it easier to trace each aligned output back to the original input locus.

## Inspecting aligned outputs
After running `alignSeqs()`, it is good practice to inspect the resulting alignment files before using them in downstream workflows.

Questions to check include:

- are all expected sequences present?
- do sequence labels look correct and consistent?
- are the alignments the expected length?
- were the files written in the intended format?

Once the outputs are confirmed, they can be loaded into R for concatenation.

## Loading aligned files into R
If aligned outputs were written in `NEXUS` format, they can be loaded directly with `ape::read.nexus.data()`.

```{r, eval=FALSE}
genes <- list.files("RESULTS_alignSeqs/09Mar2026")
my_alignments <- list()

for (i in genes) {
  my_alignments[[i]] <- ape::read.nexus.data(
    paste0("RESULTS_alignSeqs/09Mar2026/", i)
  )
}

names(my_alignments) <- gsub("[.].*", "", names(my_alignments))
```

This creates the named list structure expected by catGenes concatenation functions.

## Typical workflow after alignment
Once the alignments are loaded into R, the next step is usually:

- `catfullGenes()` for datasets with one sequence per species per locus
- `catmultGenes()` for datasets containing duplicated taxa or multiple accessions

For example:
```{r, eval=FALSE}
catdf <- catfullGenes(
  my_alignments,
  shortaxlabel = TRUE,
  missdata = TRUE
)
```

or
```{r, eval=FALSE}
catdf <- catmultGenes(
  my_alignments,
  maxspp = TRUE,
  shortaxlabel = TRUE,
  missdata = TRUE
)
```

## Common issues

### Input files are not in FASTA format
`alignSeqs()` expects `FASTA` inputs. If your files are currently in another format, convert them first or ensure the correct file type is used before alignment.

### Alignment method is not specified correctly
The method argument must match one of the supported algorithms, such as `ClustalW` or `Muscle`.

### Sequence labels are inconsistent
Alignment can still run when labels are inconsistent, but later concatenation may fail or produce unexpected results. Standardize labels before or immediately after retrieval.

### Duplicate sequences are unintentionally combined
Because `combineFASTA()` keeps all sequences, duplicated entries may persist into the alignment stage. Inspect combined files when this matters for downstream analysis.

### Output format is not appropriate for downstream use
If your next step is concatenation with `catGenes`, `NEXUS` is often the most convenient output format.

## Recommended practice
For a smooth workflow:

- combine `FASTA` files only when merging batches of the *same* locus is necessary — never mix different genes into one combined file
- align each locus as a separate dataset
- use clear and stable file names
- inspect combined and aligned outputs before concatenation
- write alignments in `NEXUS` format if you plan to load them directly into `catGenes`

## Next step
Once your sequences have been combined and aligned, the next step is usually to:

- load the alignments into R
- choose `catfullGenes()` or `catmultGenes()`
- concatenate loci for downstream phylogenetic analysis

See the next tutorials for loading alignments and concatenating multilocus datasets with `catGenes`.

