---
title: "Fish and rename taxa across multiple alignments"
format:
  html:
    toc: true
    toc-depth: 3
---

## Overview

It is common to receive or download a large, published multi-locus dataset (a supermatrix of many per-gene alignments) and only need a handful of taxa from it, for example to build a small focused analysis around a single genus or clade. Pulling those sequences out one file at a time, and then manually deleting alignment sites that become entirely empty once the other taxa are removed, is tedious and error-prone across dozens or hundreds of loci.

`catGenes` provides two companion functions for this:

- `fishSeqs()`: keeps only the sequences you ask for across every alignment in a folder, and removes any site (column) left entirely composed of gaps ("-") or missing data ("?") once the other taxa are gone.
- `renameTaxa()`: relabels sequences across a folder of alignments using a lookup table, leaving any unmatched name unchanged.

Both work the same way as other `catGenes` batch functions such as `convertAlign()` and `alignSeqs()`: point them at a directory containing NEXUS, FASTA or PHYLIP alignments (any mix is fine), and they process every file in it.

## When to use `fishSeqs()`

Use `fishSeqs()` when:

- you have a large set of per-locus alignments (a published supermatrix, for instance) and only need a subset of taxa from it
- you want each resulting alignment to remain valid, i.e. without leftover all-gap columns from the taxa you removed
- you want to process many loci in a folder at once instead of editing each file by hand

## Basic usage

```{r, eval=FALSE}
library(catGenes)

fishSeqs(
  filepath = "Loci_alignment_ZHANGetal2020/coding",
  taxa = c("Moldenhawera", "Tachigali")
)
```

This reads every alignment file in `coding/`, keeps only the sequences whose names contain `"Moldenhawera"` or `"Tachigali"`, and writes the fished-out alignments to a dated subfolder under `RESULTS_fishSeqs/`.

`taxa` is matched case-insensitively as a pattern against each sequence's name, so it works directly whenever tip labels follow the usual `Genus_species_Voucher` convention used throughout `catGenes`.

## Gap-only columns are trimmed after filtering, not before

A site that carried real data for a taxon you are removing can become entirely empty once that taxon is gone. `fishSeqs()` only removes a column if it is **entirely** gaps/missing data **among the taxa you kept** — trimming always happens after filtering, so no informative site is ever discarded by mistake.

## Working with datasets that use coded labels

Not every published dataset uses `Genus_species_Voucher` tip labels. The Zhang et al. (2020) plastome dataset, for example, uses opaque voucher codes as headers (`>FC520`, `>FC459`, ...), with the genus/species information kept in a separate supplementary table. `fishSeqs()` matches directly against whatever text is in the alignment, so for a dataset like this you resolve the taxon names to their codes yourself first, then pass the codes in:

```{r, eval=FALSE}
tbl <- openxlsx::read.xlsx("Loci_alignment_ZHANGetal2020/Supplementary_Table_S1.xlsx")
codes <- tbl$Label[tbl$Genus %in% c("Moldenhawera", "Tachigali")]
# codes -> "FC520" "FC459"

fishSeqs(
  filepath = "Loci_alignment_ZHANGetal2020/coding",
  taxa = codes
)
```

## Renaming taxa with `renameTaxa()`

Once you have fished out a coded dataset like the one above, you will usually want to replace the opaque codes with something readable, for instance combining the taxon name with the original code:

```{r, eval=FALSE}
lookup <- c(
  FC520 = "Moldenhawera_blanchetiana_FC520",
  FC459 = "Tachigali_costaricensis_FC459"
)

renameTaxa(
  filepath = "RESULTS_fishSeqs/01Sep2026",
  lookup = lookup
)
```

`lookup` can also be a two-column data.frame, with the original names in the first column and the new names in the second — useful when you build it directly from a spreadsheet:

```{r, eval=FALSE}
lookup <- data.frame(
  old = tbl$Label,
  new = paste(tbl$Genus, tbl$Species, tbl$Label, sep = "_")
)

renameTaxa(
  filepath = "RESULTS_fishSeqs/01Sep2026",
  lookup = lookup
)
```

Any sequence name **not** found in `lookup` is left unchanged, so you can rename just a subset of taxa without affecting the rest of the alignment.

### Saving elsewhere vs. overwriting in place

By default, `renameTaxa()` writes the renamed alignments to a new dated subfolder under `dir` (`RESULTS_renameTaxa/` by default), leaving your original files untouched:

```{r, eval=FALSE}
renameTaxa(
  filepath = "RESULTS_fishSeqs/01Sep2026",
  lookup = lookup,
  dir = "RESULTS_renameTaxa"
)
```

If you are confident in the lookup table and want to relabel the files in place instead, set `overwrite = TRUE`:

```{r, eval=FALSE}
renameTaxa(
  filepath = "RESULTS_fishSeqs/01Sep2026",
  lookup = lookup,
  overwrite = TRUE
)
```

When `overwrite = TRUE`, each file is always rewritten in its original format; the `format` argument only applies to the default (non-overwrite) mode.

## Fishing and renaming in a single step

`fishSeqs()` accepts the same kind of lookup directly through its `rename` argument, so you can fish and relabel in one call instead of chaining two:

```{r, eval=FALSE}
fishSeqs(
  filepath = "Loci_alignment_ZHANGetal2020/coding",
  taxa = c("FC520", "FC459"),
  rename = c(
    FC520 = "Moldenhawera_blanchetiana_FC520",
    FC459 = "Tachigali_costaricensis_FC459"
  ),
  dir = "RESULTS_fishSeqs"
)
```

Renaming is applied after taxa are kept and gap-only columns are removed, so the lookup keys should match the *original* sequence names, not the trimmed alignment content.

## Forcing an output format

Both functions keep each output file in the same format as its original input file by default. Use `format` to force a common output format instead:

```{r, eval=FALSE}
fishSeqs(
  filepath = "Loci_alignment_ZHANGetal2020/coding",
  taxa = c("FC520", "FC459"),
  format = "NEXUS"
)
```

## Working across a whole folder of loci

Both functions process every file in `filepath`, so a full multi-locus dataset can be fished (and renamed) in one call:

```{r, eval=FALSE}
fishSeqs(
  filepath = "Loci_alignment_ZHANGetal2020/coding",
  taxa = codes,
  rename = lookup,
  dir = "RESULTS_fishSeqs"
)

fishSeqs(
  filepath = "Loci_alignment_ZHANGetal2020/noncoding",
  taxa = codes,
  rename = lookup,
  dir = "RESULTS_fishSeqs"
)
```

A locus that has none of the requested taxa is skipped (with a message when `verbose = TRUE`), so it is normal for the two calls above to produce a different number of output files than there are input files, especially for `noncoding` regions that were not sequenced for every accession.

## Common issues

### No taxa matched in a given file
If none of the sequence names in a file match `taxa`, that file is skipped and a warning is issued once at the end if no file produced any output at all. This is expected for loci where a given accession was not sequenced.

### Sequences in a file are not the same length
`fishSeqs()` expects each input file to already be an alignment (equal-length sequences). If a file's sequences differ in length, it is skipped with a warning rather than silently producing a broken output.

### A renamed lookup key does not match anything
`renameTaxa()` never errors on an unmatched lookup entry; it simply leaves that sequence's name unchanged. Check the "Renamed X of Y" message (`verbose = TRUE`) if you expect more names to have matched than actually did — it usually means the original code in your alignment does not exactly match the key you used in `lookup`.

## Recommended practice

- resolve genus/species names to the codes actually used in your headers **before** calling `fishSeqs()`, using whatever crosswalk table came with the dataset
- keep the default `overwrite = FALSE` for `renameTaxa()` until you have checked the renamed output, then switch to `overwrite = TRUE` if you want to relabel in place
- run with `verbose = TRUE` the first time on a new dataset so you can see per-file match counts and catch typos in `taxa`/`lookup` early

## Next step

Once you have a folder of fished (and renamed) per-locus alignments, the usual next step is to load them into R and concatenate them:

```{r, eval=FALSE}
genes <- list.files("RESULTS_fishSeqs/01Sep2026")
my_alignments <- list()

for (i in genes) {
  my_alignments[[i]] <- ape::read.nexus.data(
    paste0("RESULTS_fishSeqs/01Sep2026/", i)
  )
}

names(my_alignments) <- gsub("[.].*", "", names(my_alignments))

catdf <- catfullGenes(
  my_alignments,
  shortaxlabel = TRUE,
  missdata = TRUE
)
```

See the tutorials on concatenating multilocus datasets for the next steps.
