---
title: "Loading multiple alignments into R"
format:
  html:
    toc: true
    toc-depth: 3
---

## Overview

Most `catGenes` workflows begin with a set of individual DNA alignments imported into R as a named list. This article explains how to load multiple alignments, organize them consistently, and prepare them for downstream functions such as `catfullGenes()` and `catmultGenes()`.

The examples below use alignments in **NEXUS format** read with `ape::read.nexus.data()`, which is the format most commonly used in `catGenes` concatenation workflows.

---

## Before you start

Before loading alignments into R, make sure that:

- taxon labels are consistently formatted across loci  
- file names are informative and stable  
- each alignment corresponds to a single gene or locus  
- all input files are stored in the same folder if you want to import them together  

For guidance on sequence label formatting, see the article on **standardizing DNA alignments for `catGenes`**.

---

# Loading example alignments included in `catGenes`

The package includes example alignments for **Vataireoid legumes**. These files can be loaded directly from the installed package and are useful for learning the basic workflow.

```{r eval=FALSE}

library(catGenes)

genes <- list.files(system.file("DNAlignments/Vataireoids",
                                package = "catGenes"))

Vataireoids <- list()

for (i in genes) {
  Vataireoids[[i]] <- ape::read.nexus.data(
    system.file("DNAlignments/Vataireoids", i, package = "catGenes")
  )
}

names(Vataireoids) <- gsub("[.].*", "", names(Vataireoids))
```

The resulting object is a named list of DNA alignments.

```{r eval=FALSE}

names(Vataireoids)
```

Each element of the list corresponds to a single alignment, and the list names indicate the gene or locus.

# Loading alignments from your own directory

If your DNA alignments are stored in a folder on your computer, you can import them using `list.files()` and a loop.
```{r}
#| echo: false

library(catGenes)

genes <- list.files("path_to_DNA_alignments_folder")
my_alignments <- list()

for (i in genes) {
  my_alignments[[i]] <- ape::read.nexus.data(
    paste0("path_to_DNA_alignments_folder/", i)
  )
}
```


# Recommended file naming

For simplicity, alignment files should ideally be named only by locus, for example:

```{r eval=FALSE}
ITS.nex
matK.nex
rbcL.nex
trnDT.nex
```

This keeps the resulting list names easy to interpret and helps preserve clear partition names in downstream concatenation workflows.

# Loading files with shared prefixes
In some datasets, alignment files may include a dataset prefix before the gene name, such as:

```{r eval=FALSE}
Vataireoids_ITS.nex
Vataireoids_matK.nex
Vataireoids_trnDT.nex
```
These files can still be loaded normally, and the list names can then be cleaned to retain only the locus name.

```{r eval=FALSE}
library(catGenes)

genes <- list.files("path_to_DNA_alignments_folder")
my_alignments <- list()

for (i in genes[grepl("Vataireoids", genes)]) {
  my_alignments[[i]] <- ape::read.nexus.data(
    paste0("path_to_DNA_alignments_folder/", i)
  )
}

names(my_alignments) <- gsub(".*_(.+)[.].*", "\\1", names(my_alignments))
```

This approach is useful when several projects or clades are stored in the same directory.


# Loading alignments one by one
If you prefer, you can import each alignment separately and then combine them into a list.

```{r eval=FALSE}
library(catGenes)

ITS   <- ape::read.nexus.data("ITS.nex")
matK  <- ape::read.nexus.data("matK.nex")
trnDT <- ape::read.nexus.data("trnDT.nex")

my_alignments <- list(
  ITS = ITS,
  matK = matK,
  trnDT = trnDT
)
```
This approach is practical for small datasets or when you want more explicit control over the list structure.

## Common issues

# File paths are incorrect
If R cannot find your files, check that the path provided to `list.files()` or `ape::read.nexus.data()` is correct.

# Files are not in NEXUS format
The examples in this article assume NEXUS input. If your files are in another format, you may need to convert them first using `convertAlign()`.

# List names are not informative
If you do not remove file extensions or prefixes, the resulting list names may be difficult to interpret in downstream analyses. Clean list names as early as possible.

# Taxa do not match across loci
If concatenation behaves unexpectedly, the problem is often due to inconsistent taxon labels rather than the import step itself. In that case, revisit the alignment standardization tutorial.

## Next step
Once your alignments are loaded into R as a named list, the next step is to choose the appropriate concatenation workflow:

- use `catfullGenes()` when each species has a single sequence per locus
- use `catmultGenes()` when one or more species have multiple accessions

These workflows are described in the dedicated tutorials on concatenating multilocus datasets with `catGenes`.
