---
title: "Model selection and MrBayes preparation"
format:
  html:
    toc: true
    toc-depth: 3
---

## Overview

After concatenating multilocus alignments and exporting the final dataset, a common next step is to determine suitable substitution models for each partition and prepare the dataset for Bayesian phylogenetic analysis. In `catGenes`, this workflow is supported by `evomodelTest()`.

The function `evomodelTest()` performs evolutionary model selection using `phangorn::modelTest()`, identifies the best-fitting model according to an information criterion, and can generate `MrBayes` command blocks for downstream analysis. When desired, it can also append these commands directly to `NEXUS` files.

This article explains how to use `evomodelTest()` for model selection and `MrBayes` preparation, how its main arguments affect the output, and how it fits into a broader `catGenes` phylogenetic workflow.

## When to use `evomodelTest()`

Use `evomodelTest()` when:

- you already have one or more aligned `NEXUS` files
- you want to compare nucleotide substitution models for each locus
- you want to select models according to `AIC` or `BIC`
- you want to generate a `MrBayes` block automatically
- you want to prepare a combined partitioned dataset for Bayesian analysis

A typical point to use this function is after:

1. aligning loci
2. loading them into R
3. concatenating the dataset
4. exporting the aligned partitions or final concatenated matrix as `NEXUS`

## Relationship with the broader workflow

A common `catGenes` workflow is:

1. retrieve or mine sequences
2. align loci
3. concatenate the dataset with `catfullGenes()` or `catmultGenes()`
4. export the final matrix with `writeNexus()`
5. use `evomodelTest()` to select models and prepare a `MrBayes` block
6. run `MrBayes` with `mrbayesRun()`

This means `evomodelTest()` serves as the bridge between dataset preparation and Bayesian phylogenetic inference.

## Basic usage

A simple model-selection workflow looks like this:

```{r, eval=FALSE}
library(catGenes)

evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  model_criteria = "BIC"
)
```


This evaluates substitution models for the provided NEXUS alignment and generates a report of model-selection results.

## Required input

The main input is nexus_file_path, which points to one or more NEXUS alignment files.

For example:
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = "path/to/ITS.nex"
)
```

or, for multiple locus files:
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = c("path/to/ITS.nex",
                      "path/to/matK.nex",
                      "path/to/rbcL.nex")
)
```

These can be individual alignments that you later want to use in a partitioned Bayesian analysis.

## Choosing the information criterion
The argument model_criteria controls how the best model is chosen.
Supported options include:
- "AIC"
- "BIC"

For example, using AIC:
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  model_criteria = "AIC"
)
```

In many phylogenetic workflows, BIC is a conservative and commonly preferred criterion, but both are useful depending on analytical preference.

## Choosing which models to test
The argument models_to_test controls the set of substitution models included in the comparison.
You can test all available models:

```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  models_to_test = "all"
)
```

Or a standard subset:
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  models_to_test = "standard"
)
```

You can also specify a custom vector of models:
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  models_to_test = c("JC", "HKY", "GTR")
)
```
This is useful when you want to restrict testing to a smaller, interpretable set of candidate models.

## Including gamma rate heterogeneity
The argument `include_G` controls whether among-site rate variation modeled with gamma-distributed rates should be evaluated.
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  include_G = TRUE
)
```

Including gamma-distributed rate heterogeneity is often appropriate for empirical DNA datasets.

## Including invariant sites
The argument `include_I` controls whether models including a proportion of invariant sites are tested.
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  include_I = TRUE
)
```

The combination of +G and +I options can have a substantial effect on model choice, so it is useful to make these decisions explicitly.

## Parallel computation
The argument mc.cores controls the number of cores used during model testing.
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  mc.cores = 4
)
```

Using multiple cores can speed up model testing, especially when working with multiple loci or larger alignments.

## Appending a MrBayes block to NEXUS files
One of the main strengths of `evomodelTest()` is that it can automatically generate and append MrBayes commands to the input `NEXUS` file.
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  append_mrbayes_to_nexus = TRUE
)
```

This is convenient because it prepares the alignment for direct downstream use with `MrBayes`.
If you do not want to append the block:
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  append_mrbayes_to_nexus = FALSE
)
```

## Overwriting the original NEXUS file
The argument overwrite_original_nexus controls whether the original `NEXUS` file should be replaced when the `MrBayes` block is appended.
To overwrite the original file:
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  append_mrbayes_to_nexus = TRUE,
  overwrite_original_nexus = TRUE
)
```

To preserve the original file and write a modified copy instead:
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  append_mrbayes_to_nexus = TRUE,
  overwrite_original_nexus = FALSE
)
```

Keeping the original file unchanged is often safer, especially when testing alternative model-selection settings.

## Preparing a combined partitioned MrBayes block
If you already have a concatenated `NEXUS` matrix and want to append a complete partitioned `MrBayes` block to it, use `combined_nexus_file_path`.
For example:
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = c("ITS.nex", "matK.nex", "rbcL.nex"),
  combined_nexus_file_path = "combined_dataset.nex",
  append_mrbayes_to_nexus = TRUE
)
```
This allows model selection to be performed on the individual loci while the final combined `MrBayes` block is added to the concatenated partitioned matrix.

This is one of the most useful workflows when moving from locus-specific alignments to a partitioned Bayesian analysis.

## Choosing the output directory
The argument `dir` controls where model-selection results are written.
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  dir = "RESULTS_evomodelTest"
)
```
By default, the function creates a results folder with a date-stamped subdirectory. This helps keep multiple model-selection runs organized.

## Verbose output
The argument `verbose` controls whether progress messages are printed during execution.
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  verbose = TRUE
)
```

To suppress progress messages:
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  verbose = FALSE
)
```

This can be useful when rendering tutorials or running batch workflows.

## A complete example workflow
A typical workflow might look like this:

Step 1. Export a concatenated NEXUS dataset
```{r, eval=FALSE}
writeNexus(
  catdf,
  file = "Vataireoids.nex",
  genomics = FALSE,
  interleave = TRUE,
  bayesblock = FALSE
)
```

Step 2. Run model selection and append a `MrBayes` block
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  model_criteria = "BIC",
  models_to_test = "standard",
  include_G = TRUE,
  include_I = TRUE,
  append_mrbayes_to_nexus = TRUE,
  overwrite_original_nexus = FALSE,
  dir = "RESULTS_evomodelTest"
)
```

At this point, you have:
- a model-selection report
- a `NEXUS` file with a generated `MrBayes` block
- a prepared input for downstream Bayesian analysis

## Typical workflow for partitioned multilocus analyses
A particularly useful workflow is:
- align loci separately
- export each locus as `NEXUS`
- export a concatenated partitioned matrix with `writeNexus()`
- run `evomodelTest()` on the separate loci
- append the partitioned MrBayes block to the concatenated `NEXUS` file
- run the resulting file with `mrbayesRun()`

This keeps model choice tied to each locus while preparing a combined Bayesian analysis.

The results directory may also contain:
- a detailed text report
- modified `NEXUS` files with appended MrBayes blocks
- summaries of the selected models

Inspecting these outputs is useful before proceeding to Bayesian inference.

## Common issues
### Input file is not in NEXUS format
`evomodelTest()` expects `NEXUS` alignment files. If your data are in another format, convert them first using convertAlign() or export them with w`riteNexus()`.

### Model comparison takes too long
Large alignments or many candidate models can increase runtime. To reduce computation time, you can:
- reduce the number of models tested
- lower retmax-style dataset size earlier in the workflow
- increase mc.cores
- use "standard" rather than "all"

### Original `NEXUS` file is overwritten unintentionally
Be cautious when using `overwrite_original_nexus = TRUE`. If you want to preserve the original matrix, keep this option set to `FALSE`.

### The `MrBayes` block is not appended where expected
Make sure:
- `append_mrbayes_to_nexus = TRUE`
- the `nexus_file_path` is correct
- `combined_nexus_file_path` is provided when working with a separate concatenated matrix

### The selected model does not match expectations
Different criteria (AIC vs BIC) or different choices for `include_G` and `include_I` can change the best-fitting model. This is normal and should be interpreted in the context of the dataset and analytical goals.

## Recommended practice
For the smoothest model-selection workflow:
- work from clean `NEXUS` alignments
- decide in advance whether to use AIC or BIC
- test a manageable but meaningful model set
=
- include +G and +I only when justified by the workflow
- preserve original `NEXUS` files unless replacement is intentional
- inspect the generated `MrBayes` block before starting the analysis

## Next step
Once model selection has been completed and the `MrBayes` block has been prepared, the next step is to run the Bayesian analysis itself using `mrbayesRun()` or an external `MrBayes` workflow.
