---
title: "Plot and edit phylogenetic trees"
format:
  html:
    toc: true
    toc-depth: 3
---

## Overview

The function `plotPhylo()` provides a flexible workflow for plotting and editing phylogenetic trees in `catGenes` using `ggtree`. It was designed to help users move from a raw phylogeny to a more informative and publication-ready figure, with options for changing layouts, displaying support values, highlighting clades or taxa, modifying labels, pruning tips, and exporting the final result.

This article explains how to use `plotPhylo()` in a range of common situations, from simple tree plotting to more customized phylogenetic figures.

## When to use `plotPhylo()`

Use `plotPhylo()` when you want to:

- visualize a phylogenetic tree in R
- add node support information
- highlight particular clades or taxa
- rename or prune selected tips
- adjust tip label formatting
- create more presentation-ready or publication-ready figures
- save the resulting tree as an image file

The function is especially useful after a phylogenetic analysis has already been completed and the resulting tree is available as a `phylo` object.

## Required input

The main required input is a tree of class `phylo`.

For example:
```{r, eval=FALSE}
library(catGenes)
library(ape)

tree <- read.tree("my_tree.tre")
```

Once the tree has been imported, it can be passed to `plotPhylo()`.

## Basic usage
A minimal example looks like this:
```{r, eval=FALSE}
p <- plotPhylo(tree = tree)
```

This returns a ggtree-based plot object that can be printed, modified, or saved.

## Choosing the tree layout

The `layout` argument controls how the tree is displayed. Supported layouts include:

- "rectangular"
- "dendrogram"
- "slanted"
- "ellipse"
- "roundrect"
- "fan"
- "circular"
- "inward_circular"
- "radial"
- "equal_angle"
- "daylight"
- "ape"

A simple rectangular layout:
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  layout = "rectangular"
)
```

A circular layout:
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  layout = "circular"
)
```

The best layout depends on the size of the tree and the intended figure style.

## Controlling branch width
The `branch.width` argument sets the line width of tree branches.
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  branch.width = 0.5
)
```

Thicker branches can be helpful in large figures or presentations, whereas thinner branches are often preferable in complex or crowded trees.

## Showing or hiding branch supports
The `branch.supports` argument controls whether node support values are displayed.
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  branch.supports = TRUE
)
```

This is useful when support values are stored in the tree object and you want them rendered automatically.

## Adding support from alternative trees
`plotPhylo()` can also display support values from other phylogenetic analyses, using alternative tree objects.

Add support from a RAxML tree
```{r, eval=FALSE}
p <- plotPhylo(
  tree = main_tree,
  add.raxml.tree = raxml_tree
)
```

Add support from a parsimony tree
```{r, eval=FALSE}
p <- plotPhylo(
  tree = main_tree,
  add.parsi.tree = parsimony_tree
)
```
This allows support from different inference methods to be compared and visualized together on the same phylogeny.

## Showing or hiding tip labels
The `show.tip.label` argument controls whether the tip labels are displayed.
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  show.tip.label = TRUE
)
```

Hiding tip labels may be useful when creating an overview figure or focusing on clade structure rather than taxon names.

## Adjusting tip label size
Use `size.tip.label` to change the font size of tip labels.
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  size.tip.label = 2
)
```

Smaller values help when plotting large trees, while larger values can be useful for smaller trees or presentation figures.

## Changing the font face of tip labels
Use fontface.tip.label to control the text style of the tip labels.
Common options include:
```{r, eval=FALSE}
"plain"
"bold"
"italic"
"bold.italic"
```

For example:
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  fontface.tip.label = "italic"
)
```

Italic labels are often appropriate for scientific names.

# Using fancy tip labels
The `fancy.tip.label` argument allows distinct formatting for taxon names and associated identifiers such as voucher codes or accession numbers.
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  fancy.tip.label = TRUE
)
```

This is especially useful when the tree labels contain both taxonomic and accession-level information and you want a more informative presentation.

## Abbreviating tip labels
The `abbrev.tip.label` argument can be used when tip labels are long and you want a more compact display.
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  abbrev.tip.label = TRUE
)
```
This can make dense trees easier to read.

## Highlighting selected taxa
You can highlight specific taxa using highlight.taxa.
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  highlight.taxa = c("Vatairea_fusca", "Vatairea_guianensis")
)
```

You can also specify colors explicitly with highlight.color:
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  highlight.taxa = c("Vatairea_fusca", "Vatairea_guianensis"),
  highlight.color = c("red", "blue")
)
```

This is useful when emphasizing focal taxa in a large tree.

## Understating selected taxa
The `understate.taxa` argument can be used to make some taxa visually less prominent, usually by rendering them in gray.
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  understate.taxa = c("Outgroup_species", "Outgroup_species_2")
)
```

This is helpful when you want to emphasize a focal clade while keeping the rest of the tree visible but less dominant.

## Highlighting clades
The `highlight.clade` argument allows you to highlight the clade defined by two tip labels.
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  highlight.clade = c("Species_A", "Species_B")
)
```

You can also highlight multiple clades by providing a list:
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  highlight.clade = list(
    c("Species_A", "Species_B"),
    c("Species_C", "Species_D")
  )
)
```

To control the highlighting colors, use fill.gradient.
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  highlight.clade = list(
    c("Species_A", "Species_B"),
    c("Species_C", "Species_D")
  ),
  fill.gradient = c("lightblue", "mistyrose")
)
```

This is useful when marking major groups or focal clades in a figure.

## Renaming tip labels
The `replace.taxa` argument lets you substitute specific tip labels with alternative names.
For example:
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  replace.taxa = c(
    "Harpalyce_formosa" = "Harpalyce_riparia",
    "Harpalyce_cf_brasiliana" = "Harpalyce_magnibracteata"
  )
)
```

This is useful when updating names, correcting labels, or standardizing terminology in the plotted figure without altering the original tree object.

## Pruning selected taxa
The `prune.taxa` argument allows you to drop selected tips from the plotted tree.
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  prune.taxa = c("Outgroup_species", "Taxon_to_remove")
)
```
This can be useful when producing a focused figure that excludes outgroups or taxa not relevant to the current presentation.

## Adding a gene or dataset label
The `gene.label` argument adds a label or title associated with the plotted tree.
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  gene.label = "matK"
)
```

This is useful when comparing trees built from different loci or different datasets.
You can also adjust its placement with ylim.gene.label.

## Adjusting the x-axis
Use `xlim.tree` or `hexpand` to control horizontal spacing and tree extent.
For example:
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  xlim.tree = c(0, 1.5)
)
```

or
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  hexpand = 0.1
)
```

These options are especially useful when labels are long and need extra space.

## Adding a side phylogram
```{r, eval=FALSE}
plotPhylo() can add a small side phylogram showing branch lengths.
p <- plotPhylo(
  tree = tree,
  phylogram.side = TRUE
)
```

To show support values in the side phylogram as well:
```{r, eval=FALSE}
p <- plotPhylo(
  tree = tree,
  phylogram.side = TRUE,
  phylogram.supports = TRUE
)
```
You can also adjust its size with phylogram.height.
This is useful when the main edited tree emphasizes labels and clades, while the side phylogram preserves branch-length context.

## Saving the figure
You can save the plotted tree directly to disk using `save = TRUE`.
```{r, eval=FALSE}
plotPhylo(
  tree = tree,
  layout = "rectangular",
  save = TRUE,
  dir = "RESULTS_edited_tree",
  filename = "my_tree",
  format = "pdf"
)
```

Supported output formats include:

- "pdf"
- "jpg"
- "png"
- "tiff"

For example, to save as PNG:
```{r, eval=FALSE}
plotPhylo(
  tree = tree,
  save = TRUE,
  filename = "my_tree",
  format = "png"
)
```

## Adjusting image resolution
Use the dpi argument to set output resolution.
```{r, eval=FALSE}
plotPhylo(
  tree = tree,
  save = TRUE,
  filename = "my_tree",
  format = "tiff",
  dpi = 600
)
```

Higher resolution is useful for publication-quality raster outputs.

## A complete example
A typical tree-plotting workflow might look like this:
```{r, eval=FALSE}
library(catGenes)
library(ape)

tree <- read.tree("my_tree.tre")

p <- plotPhylo(
  tree = tree,
  layout = "rectangular",
  branch.supports = TRUE,
  show.tip.label = TRUE,
  size.tip.label = 2,
  fontface.tip.label = "italic",
  highlight.taxa = c("Vatairea_fusca", "Vatairea_guianensis"),
  highlight.color = c("red", "blue"),
  gene.label = "Combined dataset"
)
```

To save the same tree:
```{r, eval=FALSE}
plotPhylo(
  tree = tree,
  layout = "rectangular",
  branch.supports = TRUE,
  show.tip.label = TRUE,
  size.tip.label = 2,
  fontface.tip.label = "italic",
  gene.label = "Combined dataset",
  save = TRUE,
  filename = "combined_dataset_tree",
  format = "pdf"
)
```

## Typical use cases
`plotPhylo()` is especially useful for:

- final figure preparation for manuscripts
- highlighting focal taxa or clades
- comparing support values from different inference methods
- cleaning labels before presentation
- generating multiple views of the same phylogeny with different layouts

## Common issues
### Tree object is not of class phylo
`plotPhylo()` expects a tree object compatible with ape and ggtree. If the input is not a proper phylo object, convert or re-import the tree before plotting.

### Tip labels are too long
Long labels can crowd the plot. Consider:
- reducing `size.tip.label`
- using `abbrev.tip.label = TRUE`
- using `fancy.tip.label = TRUE`
- expanding the `x-axis` with `xlim.tree` or `hexpand`

### Highlighted taxa are not found
If a taxon name passed to highlight.taxa, replace.taxa, or prune.taxa does not match the exact tree label, it will not be applied. Always check the tree labels first.

### Support values are missing
If `branch.supports = TRUE` but no support values appear, the support information may not be present in the tree object or may not be formatted in the expected way.

### Saved figure does not look as expected
Try adjusting:

- layout
- size.tip.label
- xlim.tree
- hexpand
- format
- dpi
Tree figures often require some iteration to achieve the most readable final design.

## Recommended practice
For the smoothest use of `plotPhylo()`:

- start from a clean phylo object
- inspect tip labels before adding highlights or replacements
- use rectangular layouts for readability and circular or fan layouts for dense trees
- keep scientific names italicized when appropriate
- save final figures in pdf for vector output or tiff/png for high-resolution raster output
- make small adjustments iteratively until the visual balance is right

## Next step
Once your phylogeny has been plotted and edited with `plotPhylo()`, the figure can be used directly for manuscripts, presentations, reports, or supplementary materials as part of a reproducible phylogenetic workflow in R.
