---
title: "Retrieve sequences from GenBank with accession numbers"
format:
  html:
    toc: true
    toc-depth: 3
---

## Overview

The function `mineSeq()` retrieves DNA sequences from GenBank using accession numbers and returns them in a format ready for downstream use in `catGenes`. This workflow is useful when you already have a table of GenBank accessions for one or more loci and want to download the corresponding sequences in a standardized way.

Depending on the input and options used, `mineSeq()` can:

- retrieve sequences directly from GenBank
- preserve the original GenBank taxonomy
- rename sequences using your own taxon and voucher information
- return sequences as `DNAbin` objects or character strings
- save the results to disk as `FASTA` files

This article shows how to organize the input table, run `mineSeq()`, inspect the results, and prepare the output for later alignment and concatenation steps.

## When to use `mineSeq()`

Use `mineSeq()` when:

- you already know the GenBank accession numbers you want to download
- you have a spreadsheet or data frame listing accessions for one or more genes
- you want to standardize sequence names before alignment
- you want to build locus-specific or multilocus datasets from an accession table

If you do not already have accession numbers and instead want to search GenBank by taxonomic query terms, see the tutorial on retrieving sequences with `mineTaxa()`.

## Input table structure

The main input to `mineSeq()` is a data frame containing GenBank accession numbers. Ideally, this data frame includes:

- a `Species` column with taxon names
- a `Voucher` column with voucher or accession identifiers
- one column for each locus or gene containing the relevant GenBank accessions

For example:
```{r, eval=FALSE}
example_accessions <- data.frame(
  Species = c("Vatairea_fusca", "Vatairea_guianensis"),
  Voucher = c("Cardoso2939", "Silva1820"),
  ITS = c("JX152598", "JX152599"),
  matK = c("JX152610", "JX152611"),
  rbcL = c("JX152620", "JX152621"),
  stringsAsFactors = FALSE
)
```

In this structure:
- each row corresponds to a taxon or accession
- each gene column contains a GenBank accession number
- missing accessions can be represented as `NA`

## Selecting the accession columns
The argument `gb.colnames` tells `mineSeq()` which columns contain GenBank accession numbers.
For the example above:
```{r, eval=FALSE}
gb.colnames = c("ITS", "matK", "rbcL")
```

This allows the function to retrieve sequences only from the relevant columns.

## Basic usage
A typical `mineSeq()` call looks like this:
```{r, eval=FALSE}
library(catGenes)

seqs <- mineSeq(
  inputdf = example_accessions,
  gb.colnames = c("ITS", "matK", "rbcL")
)
```

This retrieves the sequences associated with the listed accessions and returns them as an R object.

## A real example with package data
If your package includes a dataset such as GenBank_accessions, you can use it directly.

```{r, eval=FALSE}
library(catGenes)

data(GenBank_accessions)

seqs <- mineSeq(
  inputdf = GenBank_accessions,
  gb.colnames = c("ETS", "ITS", "matK", "petBpetD", "trnTF", "Xdh")
)
```

This is a convenient way to test the workflow with a larger example dataset.

## How `mineSeq()` names retrieved sequences
When `Species` and `Voucher` columns are present in the input data frame, `mineSeq()` uses them to build standardized labels for the downloaded sequences.
This is especially useful for downstream workflows, because the resulting sequence names can already match the naming logic expected by alignment and concatenation functions.
For example, a sequence may be labeled as something like:

```{r, eval=FALSE}
Vatairea_fusca_Cardoso2939
```

If the `Species` and `Voucher` columns are not present, `mineSeq()` falls back to the taxonomy and description provided by GenBank.

## Saving sequences to disk
By default, mineSeq() can save the retrieved sequences to disk as a `FASTA` file.
```{r, eval=FALSE}
mineSeq(
  inputdf = example_accessions,
  gb.colnames = c("ITS", "matK", "rbcL"),
  save = TRUE,
  filename = "GenBank_seqs",
  dir = "RESULTS_mineSeq"
)
```

This creates an output directory and writes the downloaded sequences there.

## Understanding the output directory
When `save = TRUE`, `mineSeq()` creates a directory structure under the folder specified by dir, usually including a date-based subfolder. This helps keep sequence downloads organized across runs.
For example, the output may be saved in a location such as:

```{r, eval=FALSE}
RESULTS_mineSeq/09Mar2026/
```

with a file like:
```{r, eval=FALSE}
GenBank_seqs.fasta
```

This output can then be used directly in later steps such as combining `FASTA` files or running multiple sequence alignment.

## Working with missing accessions
In many real datasets, some taxa will lack accessions for one or more loci. In those cases, it is fine to keep `NA` values in the input table.

For example:
```{r, eval=FALSE}
example_accessions_missing <- data.frame(
  Species = c("Vatairea_fusca", "Vatairea_guianensis"),
  Voucher = c("Cardoso2939", "Silva1820"),
  ITS = c("JX152598", NA),
  matK = c("JX152610", "JX152611"),
  rbcL = c(NA, "JX152621"),
  stringsAsFactors = FALSE
)
```

`mineSeq()` will attempt to retrieve only the accession numbers that are actually present.

## Preparing sequences for alignment
Once sequences have been retrieved, a common next step is to align them with `alignSeqs()`.
For example, after saving the downloaded sequences:
```{r, eval=FALSE}
alignSeqs(
  filepath = "RESULTS_mineSeq/09Mar2026",
  method = "ClustalW",
  format = "NEXUS"
)
```

This makes `mineSeq()` a natural entry point for a broader workflow that moves from GenBank accessions to aligned and concatenated datasets.

## Typical workflow after `mineSeq()`
A common sequence-processing workflow is:

- prepare a data frame of accession numbers
- retrieve sequences with `mineSeq()`
- save them as `FASTA`
- align sequences with `alignSeqs()`
- convert or standardize alignment formats if needed
- load alignments into R
- concatenate loci with `catfullGenes()` or `catmultGenes()`

## Common issues
### Accession numbers are invalid
If an accession number is incorrect, obsolete, or misspelled, GenBank retrieval may fail for that entry. Always check accession formatting in the input table.

### Gene columns are not specified correctly
If gb.colnames does not match the actual column names of the input data frame, the function will not retrieve the intended sequences.

### Species or voucher information is missing
If the Species and Voucher columns are absent, `mineSeq()` can still retrieve the sequences, but the resulting labels will depend on GenBank metadata rather than your own standardized naming scheme.

### Internet connection is unavailable
Because `mineSeq()` connects to GenBank, it requires an active internet connection.

## Recommended practice
For the smoothest downstream workflow:

- keep accession numbers in clearly named columns
- include Species and Voucher columns whenever possible
- use stable and consistent voucher identifiers
- save retrieved sequences to disk so they can be aligned later
- inspect output labels before moving to alignment or concatenation

## Next step
Once sequences have been retrieved with mineSeq(), the next step is usually to:

- if the same locus was also retrieved from another source (a separate `mineSeq()`/`mineTaxa()` run, or extracted from whole plastomes with `minePlastome()`), combine those same-locus files into one with `combineFASTA()` — see [Combining the same locus mined from different sources](/articles/combine_and_align_sequences.qmd#combining-the-same-locus-mined-from-different-sources)
- align the sequences for each locus with `alignSeqs()`
- convert the resulting alignments if needed
- load them into R for concatenation with `catGenes`

See the next tutorials for sequence alignment, format conversion, and multilocus concatenation workflows.
