---
title: "Retrieve sequences from GenBank with taxonomic queries"
format:
  html:
    toc: true
    toc-depth: 3
---

## Overview

The function `mineTaxa()` retrieves DNA sequences from GenBank using taxonomic query terms rather than precompiled accession tables. This workflow is useful when you want to search GenBank directly for sequences belonging to a taxonomic group, gene, or combination of search terms, and then download and standardize the resulting sequences for downstream use in `catGenes`.

Depending on the options used, `mineTaxa()` can:

- search GenBank using taxonomic query syntax
- retrieve large sets of sequences from the NCBI nucleotide database
- clean and standardize taxon names
- append voucher information when available
- separate complete plastomes from other sequences
- remove duplicated sequences while retaining the longest accession per species
- save the results as `FASTA` files for later alignment and concatenation

This article shows how to construct search queries, run `mineTaxa()`, understand the main arguments, and prepare the output for downstream workflows.

## When to use `mineTaxa()`

Use `mineTaxa()` when:

- you do not already have accession numbers compiled in a table
- you want to query GenBank directly by taxon name, gene, or other search terms
- you want to retrieve a large set of sequences for exploratory or comparative analyses
- you want `catGenes` to help clean taxon names and standardize output labels
- you want to separate complete plastomes from standard gene sequences

If you already have a curated table of accession numbers, the more direct function is `mineSeq()`.

## Query syntax

The main input to `mineTaxa()` is the argument `term`, which should follow NCBI Entrez search syntax. For example:

- `Leguminosae[Organism] AND matK[Gene]`
- `Arabidopsis[Organism] AND rbcL[Gene]`
- `Vatairea[Organism] AND ITS`
- `Detarioideae[Organism] AND trnL-F`

A simple example:

```{r, eval=FALSE}
mineTaxa(
  term = "Leguminosae[Organism] AND matK[Gene]"
)
```

This searches the GenBank nucleotide database for sequences annotated as belonging to Leguminosae and associated with the gene `matK`.

## Basic usage
A basic call to `mineTaxa()` looks like this:

```{r, eval=FALSE}
library(catGenes)

seqs <- mineTaxa(
  term = "Leguminosae[Organism] AND matK[Gene]"
)
```

This searches GenBank, retrieves the matching sequences, cleans the names by default, and returns the result as an R object. If `save = TRUE`, it also writes the result to disk.

## A slightly larger query
You can also request a larger number of records using retmax.
```{r, eval=FALSE}
seqs <- mineTaxa(
  term = "Leguminosae[Organism] AND matK[Gene]",
  retmax = 2000
)
```

This is useful for broad taxonomic groups or heavily represented loci.

## Choosing the database
The default database is "nucleotide", which is appropriate for most DNA sequence retrieval workflows.

```{r, eval=FALSE}
seqs <- mineTaxa(
  term = "Leguminosae[Organism] AND matK[Gene]",
  db = "nucleotide"
)
```

In principle, other NCBI databases can be specified, but the standard catGenes use case is DNA sequence retrieval from the nucleotide database.

## Cleaning taxon names
By default, `mineTaxa()` cleans and standardizes sequence labels.

```{r, eval=FALSE}
seqs <- mineTaxa(
  term = "Leguminosae[Organism] AND matK[Gene]",
  clean_taxa = TRUE
)
```

This cleaning step helps create sequence names that are easier to use in downstream alignment and concatenation workflows.

Taxon name cleaning may include:

- removing periods and hyphens
- removing or simplifying var., subsp., and similar rank indicators
- converting spaces to underscores
- trimming sp. designations
- removing trailing underscores
- simplifying hybrid cultivar formatting

This is especially useful when retrieving large numbers of GenBank records with heterogeneous metadata.

## Preserving original GenBank names
If you prefer to keep GenBank names unchanged, set `clean_taxa = FALSE`.

```{r, eval=FALSE}
seqs_raw <- mineTaxa(
  term = "Leguminosae[Organism] AND matK[Gene]",
  clean_taxa = FALSE
)
```

This preserves the original taxon labels as returned by GenBank, which can be useful for inspection or troubleshooting, but it is often less convenient for downstream concatenation.

## Adding voucher information
By default, `mineTaxa()` attempts to extract voucher information from GenBank metadata and append it to the sequence label.

```{r, eval=FALSE}
seqs <- mineTaxa(
  term = "Leguminosae[Organism] AND matK[Gene]",
  add_voucher = TRUE
)
```

When voucher information is available, this helps create more informative and stable sequence labels, for example in a format such as:
```{r, eval=FALSE}
Genus_species_voucher_accession
```

This is particularly useful when the same species is represented by multiple accessions.

# Saving the original unprocessed query

If you want to preserve a copy of the original unprocessed `FASTA` output alongside the cleaned version, use `original_query = TRUE`.

```{r, eval=FALSE}
mineTaxa(
  term = "Leguminosae[Organism] AND matK[Gene]",
  clean_taxa = TRUE,
  original_query = TRUE
)
```

This writes an additional file with the suffix _ORIGINAL_QUERY, allowing you to compare the cleaned output against the raw GenBank results.

## Separating plastomes from standard sequences

One important feature of `mineTaxa()` is the ability to separate complete plastome sequences into a separate output file.

```{r, eval=FALSE}
seqs <- mineTaxa(
  term = "Leguminosae[Organism] AND matK[Gene]",
  plastome_apart = TRUE
)
```
When this option is enabled, complete plastomes are separated into a dedicated file with a prefix such as:

```{r, eval=FALSE}
PLASTOMES_mined_seqs_by_taxon.fasta
```

This helps distinguish whole plastome accessions from ordinary gene sequences, which is often important when building clean locus-based datasets.

## Removing duplicated sequences
By default, `mineTaxa()` removes duplicate sequences from the same species, keeping only the longest one.

```{r, eval=FALSE}
seqs <- mineTaxa(
  term = "Leguminosae[Organism] AND matK[Gene]",
  rm_duplicated = TRUE
)
```

This duplicate-removal step is applied separately to plastomes and other sequences when plastome separation is active.
This behavior is useful for reducing redundancy in large GenBank downloads, especially for broad taxonomic queries.

## Keeping all retrieved sequences

If you want to retain all retrieved sequences, including duplicates, disable duplicate removal.
```{r, eval=FALSE}
seqs_all <- mineTaxa(
  term = "Leguminosae[Organism] AND matK[Gene]",
  rm_duplicated = FALSE
)
```

This can be useful when you want to inspect raw diversity in the retrieved dataset before deciding how to filter it.

## Saving results to disk
By default, `mineTaxa()` can save results to disk.
```{r, eval=FALSE}
mineTaxa(
  term = "Leguminosae[Organism] AND matK[Gene]",
  filename = "legume_matK.fasta",
  save = TRUE,
  dir = "RESULTS_mineTaxa"
)
```

This creates an output directory, usually including a date-based subfolder, and writes the retrieved sequences there.
Understanding the output directory

When `save = TRUE`, results are written to the folder specified by dir, typically under a date-stamped subdirectory. For example:
```{r, eval=FALSE}
RESULTS_mineTaxa/09Mar2026/
```

Depending on the options used, this directory may contain:

- a cleaned `FASTA` file
- an optional original-query `FASTA` file
- a separate plastome `FASTA` file
- files with duplicates removed or retained according to the settings

## Example queries
Here are some example search strategies.

### Retrieve a commonly used plastid locus for a family
```{r, eval=FALSE}
mineTaxa(
  term = "Leguminosae[Organism] AND matK[Gene]",
  retmax = 2000
)
```

### Retrieve a nuclear ribosomal marker for a genus

```{r, eval=FALSE}
mineTaxa(
  term = "Vatairea[Organism] AND ITS"
)
```

### Retrieve a broad chloroplast marker for a subfamily

```{r, eval=FALSE}
mineTaxa(
  term = "Detarioideae[Organism] AND rbcL[Gene]"
)
```

These examples can be adapted to different taxonomic groups and loci.
Typical workflow after `mineTaxa()`

A common workflow is:

- retrieve sequences from GenBank using a taxonomic query
- clean and standardize the sequence labels
- save the retrieved sequences to disk as `FASTA`
- combine or align the sequences
- convert alignment formats if needed
- load the alignments into R
- concatenate loci with `catfullGenes()` or `catmultGenes()`

For example, once the sequences have been saved:

```{r, eval=FALSE}
alignSeqs(
  filepath = "RESULTS_mineTaxa/09Mar2026",
  method = "ClustalW",
  format = "NEXUS"
)
```

This makes `mineTaxa()` a natural entry point for a broader exploratory or taxon-based workflow.

## Common issues
### Query returns too many sequences
Broad taxonomic queries may produce very large results. Refine the query by adding gene names, markers, or narrower taxonomic terms.

### Query returns unexpected records
GenBank annotations are heterogeneous. Some records may be mislabeled, incompletely annotated, or unexpectedly broad. Always inspect the output before alignment and concatenation.

### Voucher information is not available
Voucher extraction depends on GenBank metadata. Some records do not include usable voucher annotations, so labels may still rely mainly on taxon name and accession.

### Large downloads are slow
Broad queries can take time due to GenBank response limits and network latency. For very large datasets, consider splitting the retrieval into several more targeted queries.

### Cleaned taxon names require inspection
Although name cleaning is useful, it may not always behave ideally for every taxonomic group. It is good practice to inspect the cleaned output before using it in downstream workflows.

## Recommended practice
For the smoothest use of `mineTaxa()`:

- write search terms as specifically as possible
- include a gene name whenever relevant
- inspect the downloaded sequence labels before alignment
- retain voucher information when available
- save outputs to disk for reproducibility
- consider separating plastomes and removing duplicates when building standard gene datasets

## Next step
Once sequences have been retrieved with `mineTaxa()`, the next step is usually to:

- align the sequences with `alignSeqs()`
- convert the resulting alignments if needed
- load them into R
- concatenate loci for downstream phylogenetic analysis

See the next tutorials for sequence alignment, alignment conversion, and multilocus concatenation with `catGenes`.
