---
title: "Run MrBayes from R"
format:
  html:
    toc: true
    toc-depth: 3
---

## Overview

Once a concatenated `NEXUS` dataset has been prepared and a valid `MrBayes` block has been added, `catGenes` allows you to launch the Bayesian analysis directly from R using `mrbayesRun()`.

This function helps connect data preparation and phylogenetic inference in a single workflow. Rather than switching manually to a terminal session, you can start `MrBayes` from within R, direct the run to a dedicated analysis folder, and capture the program output.

This article explains how to prepare the required input, how to locate the `MrBayes` executable, how to run an analysis in foreground or background mode, and how `mrbayesRun()` fits into the broader `catGenes` workflow.

## When to use `mrbayesRun()`

Use `mrbayesRun()` when:

- you already have a `NEXUS` file with a complete and valid `MrBayes` block
- you want to launch `MrBayes` directly from R or RStudio
- you want analysis files written to a dedicated run directory
- you want standard output and error logs captured automatically
- you want a reproducible and scriptable Bayesian analysis workflow

If your `NEXUS` file does not yet contain a `MrBayes` block, first prepare it with `writeNexus()` and, if needed, `evomodelTest()`.

## What `mrbayesRun()` requires

The function requires two essential inputs:

- `nexus_file`: a `NEXUS` file that already contains a valid `MrBayes` block
- `mrbayes_dir`: the path to the directory containing the `MrBayes` executable

A minimal call looks like this:

```{r, eval=FALSE}
library(catGenes)

res <- mrbayesRun(
  nexus_file = "Vataireoids.nex",
  mrbayes_dir = "/path/to/mrbayes"
)
```


## Preparing the NEXUS file
`mrbayesRun()` does not generate the `MrBayes` block itself. The input file must already contain a valid block such as:
```{r, eval=FALSE}
begin mrbayes;
  ...
end;
```

A common workflow is:

- concatenate the dataset with `catfullGenes()` or `catmultGenes()`
- export the matrix with `writeNexus()`
- generate or append the `MrBayes` block with `evomodelTest()`
- run the resulting file with `mrbayesRun()`

For example:
```{r, eval=FALSE}
writeNexus(
  catdf,
  file = "Vataireoids.nex",
  genomics = FALSE,
  interleave = TRUE,
  bayesblock = FALSE
)

evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  model_criteria = "BIC",
  append_mrbayes_to_nexus = TRUE,
  overwrite_original_nexus = FALSE
)
```

After that, the prepared file can be passed to `mrbayesRun()`.

## Finding the MrBayes executable
The argument `mrbayes_dir` should point to the folder containing the `MrBayes` executable.
Typical examples include:

- macOS (Intel, Homebrew)
- mrbayes_dir = "/usr/local/bin"
- macOS (Apple Silicon, Homebrew)
- mrbayes_dir = "/opt/homebrew/bin"

Linux
```{r, eval=FALSE}
mrbayes_dir = path.expand("~/.local/bin")
```

Windows
```{r, eval=FALSE}
mrbayes_dir = "C:/Program Files/MrBayes"
```

If the executable is not found directly in mrbayes_dir, mrbayesRun() also checks a bin subdirectory.

## Basic usage
A standard foreground analysis looks like this:
```{r, eval=FALSE}
res <- mrbayesRun(
  nexus_file = "Vataireoids.nex",
  mrbayes_dir = "/opt/homebrew/bin"
)
```

This:

- copies or moves the `NEXUS` file into a run directory
- launches `MrBayes`
- writes output files into the run directory
- captures standard output and standard error
- returns a summary object in R

## Understanding `run_dir`
The argument `run_dir` controls where the analysis is executed.

If left as `NULL`, `mrbayesRun()` creates a new folder in the current working directory with a timestamped name.
```{r, eval=FALSE}
res <- mrbayesRun(
  nexus_file = "Vataireoids.nex",
  mrbayes_dir = "/opt/homebrew/bin",
  run_dir = NULL
)
```

You can also specify a custom directory:
```{r, eval=FALSE}
res <- mrbayesRun(
  nexus_file = "Vataireoids.nex",
  mrbayes_dir = "/opt/homebrew/bin",
  run_dir = "RESULTS_mrbayes/my_analysis"
)
```
Using an explicit run directory can be helpful when organizing multiple analyses.

## Understanding copy_mode
The argument `copy_mode` determines whether the input NEXUS file is copied or moved into the run directory before the analysis starts.

Copy the file
```{r, eval=FALSE}
res <- mrbayesRun(
  nexus_file = "Vataireoids.nex",
  mrbayes_dir = "/opt/homebrew/bin",
  copy_mode = "copy"
)
```
This is the safest and most common option because the original file is preserved.

Move the file
```{r, eval=FALSE}
res <- mrbayesRun(
  nexus_file = "Vataireoids.nex",
  mrbayes_dir = "/opt/homebrew/bin",
  copy_mode = "move"
)
```
This relocates the original file to the run directory, which may be useful in some fully automated workflows.

## Understanding `quiet`
The argument `quiet` controls whether informational messages are printed.
```{r, eval=FALSE}
res <- mrbayesRun(
  nexus_file = "Vataireoids.nex",
  mrbayes_dir = "/opt/homebrew/bin",
  quiet = FALSE
)
```

To suppress most informational output:
```{r, eval=FALSE}
res <- mrbayesRun(
  nexus_file = "Vataireoids.nex",
  mrbayes_dir = "/opt/homebrew/bin",
  quiet = TRUE
)
```
This is useful when running analyses in scripts or batch workflows.

## Understanding live
The argument `live` controls whether the MrBayes output is streamed to the R console while the analysis runs.
```{r, eval=FALSE}
res <- mrbayesRun(
  nexus_file = "Vataireoids.nex",
  mrbayes_dir = "/opt/homebrew/bin",
  live = TRUE
)
```

This is often useful for monitoring progress interactively.

To disable live console output:
```{r, eval=FALSE}
res <- mrbayesRun(
  nexus_file = "Vataireoids.nex",
  mrbayes_dir = "/opt/homebrew/bin",
  live = FALSE
)
```

## Understanding `use_processx`
When available, `mrbayesRun()` can use the processx package to improve output handling and support background execution.
```{r, eval=FALSE}
res <- mrbayesRun(
  nexus_file = "Vataireoids.nex",
  mrbayes_dir = "/opt/homebrew/bin",
  use_processx = TRUE
)
```

This is the default and generally the most robust option.

If processx is not installed or you do not want to use it:
```{r, eval=FALSE}
res <- mrbayesRun(
  nexus_file = "Vataireoids.nex",
  mrbayes_dir = "/opt/homebrew/bin",
  use_processx = FALSE
)
```

## Running in background mode
A particularly useful feature is background execution. When `background = TRUE`, `MrBayes` starts and returns control to R immediately.
```{r, eval=FALSE}
res <- mrbayesRun(
  nexus_file = "Vataireoids.nex",
  mrbayes_dir = "/opt/homebrew/bin",
  background = TRUE
)
```

In this mode, the returned object can include helper functions such as:

- poll() to check new output
- stop() to terminate the run

This is useful for long analyses.

## Monitoring a background run
If the analysis was launched with `background = TRUE`, you can inspect progress later.
For example:
```{r, eval=FALSE}
res$poll()
```
This reads new output from the running process.

# Stopping a background run
If the returned object includes a stop handle, you can terminate the run programmatically.
```{r, eval=FALSE}
res$stop()
```
This is only relevant when using background execution with `processx`.

## Understanding the output files
Because the analysis runs in the designated run directory, all standard `MrBayes` output files are written there. These may include files such as:

- parameter files
- tree files
- consensus tree files
- standard output logs
- error logs

For example, the directory may contain files like:

- analysis.nex
- analysis.nex.p
- analysis.nex.t
- analysis.nex.con.tre
- stdout.log
- stderr.log

The exact file names depend on the original input file and the `MrBayes` settings in the block.

## A complete example workflow
A full workflow from concatenation to Bayesian analysis might look like this:

Step 1. Export the concatenated dataset
```{r, eval=FALSE}
writeNexus(
  catdf,
  file = "Vataireoids.nex",
  genomics = FALSE,
  interleave = TRUE,
  bayesblock = FALSE
)
```

Step 2. Select models and append a `MrBayes` block
```{r, eval=FALSE}
evomodelTest(
  nexus_file_path = "Vataireoids.nex",
  model_criteria = "BIC",
  append_mrbayes_to_nexus = TRUE,
  overwrite_original_nexus = FALSE
)
```

Step 3. Run `MrBayes`
```{r, eval=FALSE}
res <- mrbayesRun(
  nexus_file = "Vataireoids.nex",
  mrbayes_dir = "/opt/homebrew/bin",
  copy_mode = "copy",
  live = TRUE,
  background = FALSE
)
```
This creates a fully scriptable Bayesian workflow inside R.

## Typical workflow after the run
Once the `MrBayes` analysis has finished, the next steps usually include:

- checking the parameter and log files
- examining the tree files and consensus tree
- importing the resulting phylogeny into R
- plotting and editing the tree with `plotPhylo()`

This makes `mrbayesRun()` a natural transition from data preparation to phylogenetic interpretation.

## Common issues
### MrBayes executable is not found
If the run fails immediately, the path given in `mrbayes_dir` may be incorrect. Make sure the folder really contains the `MrBayes` executable.

### The NEXUS file does not contain a valid MrBayes block
`mrbayesRun()` expects a ready-to-run NEXUS file. If the block is absent or invalid, `MrBayes` will fail. Prepare the file first with `writeNexus()` and `evomodelTest()` as needed.

### The analysis appears to hang
Long Bayesian analyses can run for substantial time, especially with large datasets. If using background mode, check progress with `poll()`.

### No output is printed to the console
If `live = FALSE`, output is not streamed interactively. Check the log files or inspect the returned object to locate stdout and stderr.

### The run directory is not where expected
If run_dir is left as `NULL`, a timestamped directory is created automatically in the working directory. If you want tighter control, specify run_dir explicitly.

### The run cannot be stopped programmatically
Programmatic stopping is mainly available when processx is installed and `background = TRUE`. In foreground mode or without processx, stopping is typically interactive.

## Recommended practice
For the smoothest use of `mrbayesRun()`:

- always prepare and inspect the `NEXUS` file first
- preserve the original file by using `copy_mode = "copy"`
- use an explicit run_dir for organized projects
- use background = TRUE for long analyses
- inspect returned paths and logs after the run
- keep the full workflow scripted for reproducibility

## Next step
Once the Bayesian analysis has completed, the next step is usually to import the resulting tree files and visualize them with `plotPhylo()` or related downstream tools.
