---
title: "Export concatenated datasets"
format:
  html:
    toc: true
    toc-depth: 3
---

## Overview

After comparing and standardizing multilocus alignments with `catfullGenes()` or `catmultGenes()`, the next step is usually to export the concatenated dataset for downstream phylogenetic analyses. In `catGenes`, this is done mainly with two functions:

- `writeNexus()` to write concatenated datasets in `NEXUS` format
- `writePhylip()` to write concatenated datasets in `PHYLIP` format and generate an associated partition file

These functions transform the list of equalized alignments returned by the concatenation workflow into files ready for downstream phylogenetic programs.

This article explains when to use each export function, how the main arguments affect the output, and how to choose between `NEXUS` and `PHYLIP` depending on the next analytical step.

## Input required by both functions

Both `writeNexus()` and `writePhylip()` expect as input the object returned by either:

- `catfullGenes()`
- `catmultGenes()`

For example, using a small example dataset:

```{r eval=FALSE}
library(catGenes)

genes <- list.files(system.file("DNAlignments/Vataireoids",
                                package = "catGenes"))

Vataireoids <- list()

for (i in genes[1:3]) {
  Vataireoids[[i]] <- ape::read.nexus.data(
    system.file("DNAlignments/Vataireoids", i, package = "catGenes")
  )
}

names(Vataireoids) <- gsub("[.].*", "", names(Vataireoids))

catdf <- catfullGenes(
  Vataireoids,
  shortaxlabel = TRUE,
  missdata = TRUE
)
```

The object catdf is now ready to be exported with either `writeNexus()` or `writePhylip()`.

## When to use writeNexus()
Use `writeNexus()` when you want:
- a concatenated dataset in `NEXUS` format
- optional interleaved or non-interleaved output
- partition definitions embedded in the matrix
- a preliminary MrBayes block with character sets
- a file structure convenient for Bayesian phylogenetic analysis and inspection

This is often the preferred output format when the next step involves `MrBayes` or when a richer, more descriptive matrix format is useful.

## When to use writePhylip()
Use writePhylip() when you want:
a concatenated dataset in `PHYLIP` format
a separate partition file describing locus boundaries
compatibility with downstream workflows that expect `PHYLIP`
a simpler matrix representation plus an external partition definition

This is especially useful when preparing datasets for software or pipelines that use `PHYLIP` together with a partition text file.

## Writing a concatenated NEXUS dataset
A basic writeNexus() workflow looks like this:
```{r eval=FALSE}
writeNexus(
  catdf,
  file = "Vataireoids.nex",
  genomics = FALSE,
  interleave = TRUE,
  bayesblock = TRUE
)
```

This writes a concatenated NEXUS file in which:
- each locus is included in the final matrix
- the matrix is interleaved
- a preliminary MrBayes block is added
- character sets are included to define the partitions

## Understanding the file argument
The argument file defines the output file name.
```{r eval=FALSE}
writeNexus(
  catdf,
  file = "my_dataset.nex"
)
```
or
```{r eval=FALSE}
writePhylip(
  catdf,
  file = "my_dataset.phy"
)
```
Use a file name that clearly reflects the dataset or project, especially when working with multiple concatenated outputs.

## Understanding genomics
Both export functions include a genomics argument, which controls how accession identifiers are preserved in the final dataset.
When `genomics = FALSE`, the output is usually more simplified and species-oriented:
```{r eval=FALSE}
writeNexus(
  catdf,
  file = "dataset.nex",
  genomics = FALSE
)
```

When `genomics = TRUE`, accession or identifier information is retained more explicitly:
```{r eval=FALSE}
writeNexus(
  catdf,
  file = "dataset.nex",
  genomics = TRUE
)
```

This is especially useful when:
- the original labels contain voucher information
- the dataset is accession-rich
- the goal is to preserve traceability between the concatenated matrix and the original sequences
In phylogenomic workflows or datasets with detailed accession information, `genomics = TRUE` is often preferable.

## Understanding interleave in writeNexus()
The argument interleave controls whether the NEXUS matrix is written as interleaved or fully concatenated.
When `interleave = TRUE`, each locus remains visually distinguished in the matrix:
```{r eval=FALSE}
writeNexus(
  catdf,
  file = "dataset.nex",
  interleave = TRUE
)
```

When `interleave = FALSE`, the dataset is written as a fully concatenated non-interleaved matrix:
```{r eval=FALSE}
writeNexus(
  catdf,
  file = "dataset.nex",
  interleave = FALSE
)
```

Interleaved output is often more readable when inspecting partitions manually, while non-interleaved output gives a more compact concatenated matrix.

## Understanding bayesblock in `writeNexus()`
The argument bayesblock controls whether a preliminary `MrBayes` block is included in the output file.
When `bayesblock = TRUE`, the output includes character sets for the partitions and a basic block structure useful for `MrBayes` workflows:
```{r eval=FALSE}
writeNexus(
  catdf,
  file = "dataset.nex",
  bayesblock = TRUE
)
```

When `bayesblock = FALSE`, the dataset is written without this block:
```{r eval=FALSE}
writeNexus(
  catdf,
  file = "dataset.nex",
  bayesblock = FALSE
)
```

This option is useful when the `NEXUS` file is needed only as a matrix or when the `MrBayes` commands will be prepared separately.

## Understanding endgaps.to.miss in writeNexus()
The argument `endgaps.to.miss` controls whether terminal gaps are converted into missing characters `(?)` in the output matrix.
```{r eval=FALSE}
writeNexus(
  catdf,
  file = "dataset.nex",
  endgaps.to.miss = TRUE
)
```

This is often desirable because terminal gaps may be more appropriately treated as missing data rather than explicit gaps in some phylogenetic workflows.
If you want to preserve terminal gaps as they are:
```{r eval=FALSE}
writeNexus(
  catdf,
  file = "dataset.nex",
  endgaps.to.miss = FALSE
)
```

## Writing a concatenated PHYLIP dataset
A basic `writePhylip()` workflow looks like this:
```{r eval=FALSE}
writePhylip(
  catdf,
  file = "Vataireoids_dataset.phy",
  genomics = FALSE,
  catalignments = TRUE,
  partitionfile = TRUE
)
```

This writes:
- a concatenated PHYLIP matrix
- a separate partition file describing the locus boundaries
This is often the preferred export route when a downstream workflow expects a simple concatenated matrix plus a separate partition definition.

## Understanding catalignments in `writePhylip()`
The argument catalignments controls whether the concatenated `PHYLIP` matrix itself is written.
```{r eval=FALSE}
writePhylip(
  catdf,
  file = "dataset.phy",
  catalignments = TRUE
)
```

In most cases, this should remain TRUE, since writing the matrix is usually the main purpose of the function.

## Understanding partitionfile in `writePhylip()`
The argument partitionfile controls whether a separate partition text file is written.
```{r eval=FALSE}
writePhylip(
  catdf,
  file = "dataset.phy",
  partitionfile = TRUE
)
```

This partition file defines the coordinate ranges for each locus in the concatenated matrix and is important for partitioned phylogenetic analyses.
If you do not need a separate partition file:
```{r eval=FALSE}
writePhylip(
  catdf,
  file = "dataset.phy",
  partitionfile = FALSE
)
```

## Understanding endgaps.to.miss in writePhylip()
As in writeNexus(), the writePhylip() function can also convert terminal gaps to missing characters.
```{r eval=FALSE}
writePhylip(
  catdf,
  file = "dataset.phy",
  endgaps.to.miss = TRUE
)
```

Or preserve them as gaps:
```{r eval=FALSE}
writePhylip(
  catdf,
  file = "dataset.phy",
  endgaps.to.miss = FALSE
)
```

## Example `NEXUS` workflow
A common `NEXUS` export workflow is:
Step 1. Load alignments
```{r eval=FALSE}
library(catGenes)

genes <- list.files(system.file("DNAlignments/Vataireoids",
                                package = "catGenes"))

Vataireoids <- list()

for (i in genes[1:3]) {
  Vataireoids[[i]] <- ape::read.nexus.data(
    system.file("DNAlignments/Vataireoids", i, package = "catGenes")
  )
}

names(Vataireoids) <- gsub("[.].*", "", names(Vataireoids))
```

Step 2. Concatenate
```{r eval=FALSE}
catdf <- catfullGenes(
  Vataireoids,
  shortaxlabel = TRUE,
  missdata = TRUE
)
```

Step 3. Write `NEXUS``
```{r eval=FALSE}
writeNexus(
  catdf,
  file = "Vataireoids.nex",
  genomics = FALSE,
  interleave = TRUE,
  bayesblock = TRUE
)
```
This produces a concatenated `NEXUS` matrix suitable for downstream Bayesian workflows.

## Example `PHYLIP` workflow
The corresponding `PHYLIP` workflow is:
Step 1. Load alignments
```{r eval=FALSE}
library(catGenes)

genes <- list.files(system.file("DNAlignments/Vataireoids",
                                package = "catGenes"))

Vataireoids <- list()

for (i in genes[1:3]) {
  Vataireoids[[i]] <- ape::read.nexus.data(
    system.file("DNAlignments/Vataireoids", i, package = "catGenes")
  )
}

names(Vataireoids) <- gsub("[.].*", "", names(Vataireoids))
```

Step 2. Concatenate
```{r eval=FALSE}
catdf <- catfullGenes(
  Vataireoids,
  shortaxlabel = TRUE,
  missdata = TRUE
)
```

Step 3. Write `PHYLIP`
```{r eval=FALSE}
writePhylip(
  catdf,
  file = "Vataireoids_dataset.phy",
  genomics = FALSE,
  catalignments = TRUE,
  partitionfile = TRUE
)
```
This produces a concatenated `PHYLIP` matrix together with a partition file.

## What the `NEXUS` output looks like
When `writeNexus()` is used with interleaving and partition definitions, the file contains:
- a concatenated `NEXUS` matrix
- each partition defined by character ranges
- optionally a preliminary MrBayes block
- The beginning of the matrix may look similar to the screenshots below:

And the end of the file may include the partition definitions:


## Keeping differing identifiers in `NEXUS` output
If you ran `catfullGenes()` or `catmultGenes()` with `shortaxlabel = FALSE`, `writeNexus()` can preserve differing identifiers across partitions while retaining a species-oriented concatenated matrix structure.
This is particularly useful when accession labels differ among loci but still need to remain traceable.

The output may appear as shown below:


## What the PHYLIP output looks like
When `writePhylip()` is used, the result usually consists of:
- a concatenated `PHYLIP` matrix
- a separate partition file
The matrix may resemble the example below:

And the partition file may look like this:


## Choosing between `NEXUS` and `PHYLIP`
Use `writeNexus()` when you want:
- a richer, self-contained matrix format
- embedded partition information
- optional `MrBayes` blocks
- a file structure convenient for Bayesian analyses and inspection

Use `writePhylip()` when you want:
- a simpler matrix format
- a separate partition file
- compatibility with external workflows expecting `PHYLIP`
In many `catGenes` projects, both exports are useful, depending on which downstream programs or analyses will be used.

## Common issues

### Output labels are not what you expected
If output labels seem too short or too detailed, check:
whether shortaxlabel was set appropriately during concatenation
whether genomics is set correctly during export

### Partition information is missing
In `writeNexus()`, make sure `bayesblock = TRUE` if you want embedded partition definitions and a preliminary `MrBayes` block.
In `writePhylip()`, make sure `partitionfile = TRUE` if you want the separate partition file.

### Terminal gaps are treated unexpectedly
Check the setting of endgaps.to.miss, especially if you need terminal gaps preserved as gaps rather than converted to missing characters.

### File names are unclear
Use output file names that clearly identify the dataset, especially when writing multiple alternative matrices with different settings.

## Recommended practice
For the smoothest export workflow:
- inspect the concatenated object before export
- decide whether identifiers should be simplified or preserved
- use `writeNexus()` for MrBayes-oriented or richly annotated outputs
- use `writePhylip()` when a matrix plus partition file is the preferred downstream input
- use clear file names for each exported dataset
- keep track of export settings in project notes or analysis scripts

## Next step
Once the concatenated dataset has been written to disk, the next step is usually to:
- select evolutionary models with `evomodelTest()`
- prepare or refine `MrBayes` blocks
- run phylogenetic analyses
- visualize resulting trees with `plotPhylo()`
